\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}

### 2014-07-03 paper/paper.Rnw
introduced figure environments
added command to interrupt/span multicol
viele mini fixes aus gemeinsamer session

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{graphicx}

@@

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Running title $\bullet$ November 2012 $\bullet$ Vol. XXI, No. 1} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
%\bibliographystyle{plain}
\bibliographystyle{unsrtnat}

@@

\title{\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing economic inequality with tax data - Switzerland from 1945 to 2010}} % Article title

>>>>

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage{float} %we need this to use figure environments within multicol. knitr figure chunks need to include fig.pos='H'. 'H' stems from package float and deactivates floating (which multicol does not like)
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{graphicx}

@@

%%%%%% new command to break multicol environment from time to time
\newcounter{tempcolnum}
\makeatletter
\newcommand{\multicolinterrupt}[1]{% Stuff to span both rows
\setcounter{tempcolnum}{\col@number}
\end{multicols}
#1%
\begin{multicols}{\value{tempcolnum}}
}
\makeatother
%%%%%%%

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\newdateformat{mydate}{\monthname[\THEMONTH] \THEYEAR}
\fancyhead[C]{Assessing income inequality with tax data $\bullet$ \mydate $\bullet$ Working paper} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
%\bibliographystyle{plain}
\bibliographystyle{unsrtnat}

@@

\title{\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}} % Article title

### 2014-07-03 paper/subdocuments/abstract.Rnw
viele mini fixes aus gemeinsamer session

\begin{abstract}
There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (\citealt{cooperation_divided_2011}, \citealt{gornick_income_2013}). 
In Switzerland, however, the development is unclear, as there is evidence for trends in both directions. Part of the inconclusive picture is due to different methodological approaches. In this paper we discuss the role of tax-data concerning the assessment of inequality in income. 
The focus of the discussion lays herein to show the benefits and shortcomings of tax data compared to current ``state of the art'' measurement concepts of economic inequality. 
We present common and new strategies to handle tax data specific methodological difficulties and compare results out of aggregated federal tax statistics to results from the Household Budget Survey (HBS). 
We can show to which extend survey data underestimate inequality in income. 
Following the results out of the tax-data Switzerland experienced in slight rise in inequality in recent years, similar to other western countries, but only because of rise in upper percentiles of the income distribution.

>>>>

\begin{abstract}
There is empirical evidence that economic inequality increased in the majority of western countries over the last decades (\citealt{oecd_divided_2011}, \citealt{gornick_income_2013}). 
In Switzerland, however, the development is unclear, as there is evidence for trends in both directions. Part of the inconclusive picture is due to different methodological approaches. In this paper we discuss the role of tax-data concerning the assessment of inequality in income. 
The focus of the discussion lays herein to show the benefits and shortcomings of tax data compared to current ``state of the art'' measurement concepts of economic inequality. 
We present common and new strategies to handle tax data specific methodological difficulties and compare results out of aggregated federal tax statistics to results from the Household Budget Survey (HBS). 
We find theoretical inconsistencies in both tax and survey data which both result in an underestimation of inequality.
Following the results from tax-data Switzerland experienced in slight rise in inequality in recent years, similar to other western countries due to a polarization of the income distribution.

### 2014-07-03 paper/subdocuments/data.Rnw
viele mini fixes aus gemeinsamer session

\section{Data, Measurement concepts and Methods}

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All these questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss the advantages and shortcomings of tax data compared to other data sources - namely survey data. We describe and explain methods and techniques needed to construct time-series of income inequality measures for Switzerland. We will present these time-series in the result section and evaluate the use of the FTA Tax Data to assess the development of inequality in Switzerland

@@

Our data comes from the Swiss Federal Tax Administration (FTA). Federal taxes are collected and documented by the FTA since 1915. Being called a war-tax in the beginning, the federal tax was renamed to crisis levy in 1934, defense-tax in 1939 and is finally known as direct federal tax since 1983. The time frame we were able to collect ranges from 1945 to 2010 including 44 tax periods for cantons and Switzerland. \footnote{Between 1993 and 2003 there is no exact data available for aggregate Switzerland (but for individual cantons) because of a system change from taxation assessed in arrears (Praenumerando-System) to taxation assessed on current year income (Postnumerando-System) which was implemented by cantons in different years.}  While the FTA provides data in electronic form since 1973 we collected earlier data by scanning hard copies. Data is available for Switzerland plus all cantons and basically covers every tax unit (individual or household) in Switzerland liable to pay federal taxes. This exempts all tax units with taxable income below a certain threshold (e.g. 15.000 CHF in 2010). Furthermore the FTA differentiates between two groups of tax units, so called normal cases and special cases. A normal case is a tax unit residing in a swiss canton without foreign source income and being liable to taxation all year long. All other tax units and very few that are taxed based on the style of living because they don't work (Pauschalbesteuerte) are special cases. \\

@@

\emph{Concepts on measuring economic resources}  \\

Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the \citet{oecd_oecd_2013} condense these ideas into the ICW framework (income, consumption and wealth), which is meant to be an internationally agreed framework on micro-level statistics.\footnote{Harmonization with other international standards was an important objective that guided the work of the expert group in developing the ICW Framework presented in this publication. Considered main standards were the System of National Accounts \citep{sna_system_2008}, the Canberra Group Handbook on Household Income Statistics \citep{united_nations_canberra_2011}, the final report of the 17th International Conference of Labour Statisticians \citep{international_labour_organisation_ilo_final_2004} and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing \citep{unece_conference_2006}}. According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recommended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or wealth holdings, or vice-versa. But it is also stated \citep[18]{oecd_oecd_2013}:"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by collecting and providing a database following this broader concept of economic well-being. \url{http://www.lisdatacenter.org/our-data/lws-database/}} \\

This last statement holds for Switzerland too, although the HBS study is strongly influenced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awareness of an assessment of income, consumption and wealth simultaneously is rising, we focus our analysis on income, which is undoubtedly a crucial indicator of economic well-being. It should be noted, that the Federal Tax Administration (FTA) publishes statistics on income and wealth but it is not possible to analyze the joint distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can in some sense be understood as mandatory consumptions.

@@

\begin{figure}
\centering
\includegraphics[width=\maxwidth]{figure/income_definition}
\caption{Income definitions}
\label{fig:incdef}
\end{figure}

@@

\emph{Statistical units} \\

The agreed standard on the statistical units, which should be the base of inequality analysis, are households not individuals \citep[60]{oecd_oecd_2013}. Indeed it is the individuals, who receive income, own assets and experience economic well-being, but their possibility to do so, is strongly tied to the concept of household. This comprises all persons under the same housing arrangement. The basic underlying assumption for collecting data on household level instead of individual level is, that people in the same household share resources and therefore pool their incomes (when two or more earners live together) and/or use the household income to provide the essentials of living for every household member (also non-earing members, like children). Additionally, there are economies of scale when people share living space and commodities and they therefore benefit from the sharing. To compare the individual economic well-being among individuals living in different households usually equivalence scales are used (see \citealt[173]{oecd_oecd_2013}, \citealt{buhmann_equivalence_1988}). \\

In tax data, however, the units are represented according to administrative rules. Tax units therefore neither represent individuals in every case nor true households. Tax units rather represent individuals and couples, but couples, who are married or officially registered. This doesn't imply, that those couples live together, as it is needed to satisfy the definition of a household. On the other hand, is it quite likely that more than one tax unit live in the same household (unmarried/unregistered couples, see \citet[99]{muller_vermogenslage_2014}). It is therefore not directly possible to elicit households and household income from tax data. This might influence the assessment of inequality development, taking into account the change from traditional household and family structures over the last century.

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. The Gini coefficient is the most known measure and mainly used for international comparison. As it is derived from the Lorenz-curve, the quantified amount of inequality can unpretentiously be described in a formal and visual way. Therefore the Gini coefficient is easily interpretable. Furthermore it has several desired statistical properties \citet{engelhardt_modelle_2000}. (1) ``\textbf{principle of population}'': the assessment of inequality is independent of the population size (2) ``\textbf{Requirement of Bresciani-Turroni}'': the measure is sensitive for changes of income shares, but not for absolute changes (e.g. doubling of all income) (3) ``\emph{weak principle of transfers}'' or ``\emph{requirement of Pigou-Dalton}'': transfers from richer households lead to a reduction of inequality. However, several drawbacks are reported in the literature. The most important point is, that the underlying distributional form of the measured inequality is unknown and it is therefore not possible to see if the measure is driven by a few rich or many poor individuals. This can also be problematic for comparison between countries or over time. In extreme cases two totally different distributions share the same Gini-coefficient.\\

@@

The recent wave of tax-data studies do not report Gini-coefficents. Rather top income shares are informed on, which are calculated not only with tax data, but together with external sources to produce the population and personal income control totals (RF 16.06.2014: hä, was? gehts da um die Randverteilung der Bevölkerung?). This procedure ensures, that the inequality measure is not biased because of non-fillers, who do not appear in tax statistics.  \citet{leigh_how_2007} compares top income shares with other inequality measures and asks, whether they are a useful measure of inequality in a society. He tries to answer this question empirically by comparing measures of inequality based on top income shares with measures of household or family inequality. He finds a strong positive relationship, but concludes (P.600): ``top income shares are far from perfect as a measure of distribution of income across society.'' Top income shares hence inform not completely on how inequality evolves elsewhere in the distribution. Furthermore, top income shares only weakly satisfy the Pigou-Dalton transfer principle (in contrast to the Gini-Coefficent as mentioned above). A transfer from rich to poor will indeed never increase the top income shares, but if the transfer is between individuals, who are either both within the top group or both outside the top group, then the share measure will remain unchanged. \\

@@

Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow better analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading). The polarizationindex is developed by \citep{Handock and Morris 1999} allows for this. Recently this index was applied in the work of \citep{alderson_exactly_2005} and \citep{gornick_how_2013}. This approach is rooted in relative distribution methods. It includes a precise comparison of the shape of two distributions (groups, over time). As the main advantage this approach allows to characterize the change in detail. It is possible to see which parts of the distribution changed, e.g. whether a polarization occurred (reduction of middle class) -- which equals an increase of inequality -- or this change is driven by either a change in the upper or lower part of the distribution.  \\

The literature mentions several other metrics with desired properties we will not discuss here (see for example \citet{cowell_chapter_2000} and \citet{hao_assessing_2010}).

@@

The issue of incomplete coverage is less dramatical with tax data. Essentially every permanent resident in Switzerland over 18 years of age (respectively 20 years of age prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). Essentially this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover (not only) foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. Most important this includes high net wealth individuals taxed according to their expenditures. Special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below 15'000 CHF and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units is reported, but not for the years before. Dell et al. (2007) try to estimate the fraction of non-taxed by comparing the reported numbers of tax units to census reports about the number of adult population. According to their estimations this fraction drops from 94\% in 1993/1994 to 63\% back in 1945/46. \\

@@

\subsection{Comparison of tax data and other data sources - advantages and shortcomings}

@@

\multicolinterrupt{
\begin{table}[H]
\caption{\label{datacomp}} 
\begin{center}
\begin{tabular}{lcc}
\hline\hline
                                   & FTA       & Survey data\\
\hline
Statistical unit                   & Tax units (individuals/married couples)  & Households\\                                   
Concepts of economic resources  & Data-driven   & Theory-driven\\
Population coverage & total population & sample \\ 
Top incomes covered                 & Yes   & No\\
Low incomes covered?               & no        & yes\\
Comparison over time possible 	   & Long       & Short \\
Cross country comparison possible? & limited   & possible\\
\hline
\end{tabular}
\end{center}
\end{table}
}
%RF: damit haben wir WER, WANN, WAS, wollen wir noch ein WOMIT/WAS DARSTELLEN -> methodik -> reldist ? jaa

To define a standard of measuring economic resources and related inequality we discussed different dimensions of demands the data needs to meet. To sum up, ideally we want to measure \emph{disposable income} for \emph{all} swiss \emph{households} yearly for an extended period of time. Tax data has one weakness when it comes to the ideal income measure as we need to work with ``taxable income'' which is designed to serve taxation purposes. However, both concepts -- what needs to be taxed and what can be spent -- are related as both address a distinction between necessary and voluntary expenses. By subtracting taxes from taxable income we are confident to satisfyingly come close the theoretically ideal income measure of disposable income. The second most important drawback with tax data is that it does not adequately address households. There are few situations in which the tax unit equals the household, that is for individuals living alone, with married partner and/or deductible children and nobody else. This ``classic'' household setup however became less common in the last decades so we need to assume statements about household inequality based on taxed data became more and more biased. The swiss law seems to be slightly outdated in this context.
Finally however, another advantage of tax data (over survey data) is indeed the observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HBS data). The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are rather consistent for the complete observation period\footnote{Limitations might however exist due to tax evasion and major changes in the tax system, e.g. 1945. Furthermore, tax units with incomes too small to quality for federal taxation are not documented before 1995 and treated as zeros afterwards making it difficult to track income development in the lower percentiles.}. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.

>>>>

\section{Data and measurement concepts}

Studies on inequality have to address several thorny challenges. It starts with answering three crucial questions: First of all one has to define, which concepts should to be looked at. This refers to answering the question about inequality of what. Secondly one has to be clear about the unit of analysis. This refers to answering the question about inequality among whom. Thirdly, one has to choose an appropriate measure of inequality. All these questions are ideally answered considering theory and a given research question. Often it has to be answered in context of a given dataset. Therefore we start this section with a description of the FTA-Tax Data. Based on a review on the literature about the measurement concepts in an ideal world, we discuss the advantages and shortcomings of tax data compared to other data sources - namely survey data.

@@

Our data comes from the Swiss Federal Tax Administration (FTA). Federal taxes are collected and documented by the FTA since 1915. Being called a war-tax in the beginning, the federal tax was renamed to crisis levy in 1934, defense-tax in 1939 and is finally known as direct federal tax since 1983. The time frame we were able to collect ranges from 1945 to 2010 including 44 tax periods for cantons and Switzerland. \footnote{Between 1993 and 2003 there is no exact data available for aggregate Switzerland (but for individual cantons) because of a system change from taxation assessed in arrears (Praenumerando-System) to taxation assessed on current year income (Postnumerando-System) which was implemented by cantons in different years.}  While the FTA provides data in electronic form since 1973 we collected earlier data by scanning hard copies. Data is available for Switzerland plus all cantons and basically covers every tax unit (individual or household) in Switzerland liable to pay federal taxes. This exempts all tax units with taxable income below a certain threshold (e.g. CHF 16.900 for singles in 2010). Furthermore the FTA differentiates between two groups of tax units, so called normal cases and special cases. A normal case is a tax unit residing in a swiss canton without foreign source income and being liable to taxation all year long. All other tax units and very few that are taxed based on the style of living because they don't work (Pauschalbesteuerte) are special cases. \\

@@

\emph{Concepts on measuring economic resources}  \\
Most studies on inequality focus on income inequality solely. However, recent activities emphasize the need of a broader conceptualization. A recent publication from the \citet{oecd_oecd_2013} condense these ideas into the ICW framework (income, consumption and wealth), which is meant to be an internationally agreed framework on micro-level statistics.\footnote{Harmonization with other international standards was an important objective that guided the work of the expert group in developing the ICW Framework presented in this publication. Considered main standards were the System of National Accounts \citep{sna_system_2008}, the Canberra Group Handbook on Household Income Statistics \citep{united_nations_canberra_2011}, the final report of the 17th International Conference of Labour Statisticians \citep{international_labour_organisation_ilo_final_2004} and the UNECE/CES recommendations for the 2010 Censuses of Population and Housing \citep{unece_conference_2006}}. According to the framework it is best to look at income, consumption and wealth as three separate but interrelated dimensions of people's economic well-being. To gain policy relevant insight, it is recommended to look at the distribution of all three distributions simultaneously. Some households with low income, for example, may report adequate levels of consumption expenditure or wealth holdings, or vice-versa. But it is also stated \citep[18]{oecd_oecd_2013}:"[...] integrated analysis at the household level has significant data requirements that go beyond the measurement efforts currently undertaken in most countries."\footnote{The Luxembourg Wealth Study Database is currently facing this shortcomings by collecting and providing a database following this broader concept of economic well-being. \url{http://www.lisdatacenter.org/our-data/lws-database/}} \\

This last statement holds for Switzerland too, although the HBS study is strongly influenced by the recommendations of the Canberra group handbook (United Nations, 2011), which concepts are part of the ICW framework. Albeit the awareness of an assessment of income, consumption and wealth simultaneously is rising, we focus our analysis on income, which is undoubtedly a crucial indicator of economic well-being. It should be noted, that the Federal Tax Administration (FTA) publishes statistics on income and wealth but it is not possible to analyze the joint distribution on the individual or household level. Also measures of consumption are largely missing in tax data, albeit deductions can in some sense be understood as mandatory consumptions.\\

@@

\multicolinterrupt{
\begin{figure}[H]
\centering
\includegraphics[width=\maxwidth]{figure/income_definition}
\caption{Income definitions}
\label{fig:incdef}
\end{figure}
}

@@

\emph{Statistical units} \\
The agreed standard on the statistical units, which should be the base of inequality analysis, are households not individuals \citep[60]{oecd_oecd_2013}. Indeed it are individuals, who receive income, own assets and experience economic well-being, but their possibility to do so, is strongly tied to the concept of household. A household comprises all persons under the same housing arrangement. The basic underlying assumption for collecting data on household level instead of individual level is, that people in the same household share resources and therefore pool their incomes (when two or more earners live together) and/or use the household income to provide the essentials of living for every household member (also non-earing members, like children). Additionally, there are economies of scale when people share living space and commodities and they therefore benefit from the sharing. To compare the individual economic well-being among individuals living in different households usually equivalence scales are used (see \citealt[173]{oecd_oecd_2013}, \citealt{buhmann_equivalence_1988}). \\

In tax data, however, the units are represented according to administrative rules. Tax units therefore neither represent individuals in every case nor true households. Tax units rather represent individuals and couples, but couples, who are married or officially registered. This doesn't imply, that those couples live together, as it is needed to satisfy the definition of a household. On the other hand, is it quite likely that more than one tax unit lives in the same household (unmarried/unregistered couples, see \citet[99]{muller_vermogenslage_2014}). It is therefore not directly possible to elicit households and household income from tax data. This might influence the assessment of inequality development, taking into account the change from traditional household and family structures over the last century. \\

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. The Gini coefficient is the most known measure and mainly used for international comparison. As it is derived from the Lorenz-curve, the quantified amount of inequality can unpretentiously be described in a formal and visual way. Therefore the Gini coefficient is easily interpretable. Furthermore it has several desired statistical properties \citet{engelhardt_modelle_2000}. (1) ``\textbf{principle of population}'': the assessment of inequality is independent of the population size (2) ``\textbf{Requirement of Bresciani-Turroni}'': the measure is sensitive for changes of income shares, but not for absolute changes (e.g. doubling of all income) (3) ``\textit{weak principle of transfers}'' or ``\textit{requirement of Pigou-Dalton}'': transfers from richer households lead to a reduction of inequality. However, several drawbacks are reported in the literature. The most important point is, that the underlying distributional form of the measured inequality is unknown and it is therefore not possible to see if the measure is driven by a few rich or many poor individuals. This can also be problematic for comparison between countries or over time. In extreme cases two totally different distributions share the same Gini-coefficient.\\

@@

The recent wave of tax-data studies do not report Gini-coefficents. Rather top income shares are informed on, which are calculated not only with tax data, but corrected using marginal distributions of income and population using external sources (census data and national accounts). This procedure ensures, that the inequality measure is not biased because of non-fillers, who do not appear in tax statistics.  \citet{leigh_how_2007} compares top income shares with other inequality measures and asks, whether they are a useful measure of inequality in a society. He tries to answer this question empirically by comparing measures of inequality based on top income shares with measures of household or family inequality. He finds a strong positive relationship, but concludes (P.600): ``top income shares are far from perfect as a measure of distribution of income across society.'' Top income shares hence inform not completely on how inequality evolves elsewhere in the distribution. Furthermore, top income shares only weakly satisfy the Pigou-Dalton transfer principle (in contrast to the Gini-Coefficent as mentioned above). A transfer from rich to poor will indeed never increase the top income shares, but if the transfer is between individuals, who are either both within the top group or both outside the top group, then the share measure will remain unchanged. \\

@@

Newer branches of inequality studies emphasize the need for broader measures of inequality, which allow more detailed analyses about the change of inequality and namely statements about the area of change (downgrading/upgrading). The polarization index developed by \citep{handcock_relative_1999} is capable of doing this. Recently this index was applied in the work of \citep{alderson_exactly_2005} and \citep{gornick_how_2013}. This approach is rooted in relative distribution methods. It includes a precise comparison of the shape of two distributions (groups, over time). As the main advantage this approach allows to characterize the change in detail. It is possible to see which parts of the distribution changed, e.g. whether a polarization occurred (reduction of middle class) -- which equals an increase of inequality -- or this change is driven by either a change in the upper or lower part of the distribution.  \\

The literature mentions several other metrics with desired properties we will not discuss here (see for example \citet{cowell_chapter_2000} and \citet{hao_assessing_2010}).\\

@@

The issue of incomplete coverage is less dramatic with tax data. Essentially every permanent resident in Switzerland over 18 years of age (respectively 20 years of age prior to 1996) is taxed on a yearly base (or every two years before the change of the tax system). Basically this leads to a full representation of the adult population of Switzerland and a complete coverage of the income distribution. This includes a separation of normal cases, which embrace the majority of taxpayers, and the special cases, which cover (not only) foreign nationals living in Switzerland but with a yearly or any other temporary resident permit only. Most important this includes high net wealth individuals taxed according to their expenditures. Special attention has to be paid to tax units with none or very low incomes. Even though they have to hand in a tax return, their income does not show up in the statistics if their income after deductions falls below the threshold (e.g. CHF 16.900 for singles in 2010) and they are therefore not taxed with direct federal taxes. This is possible for normal and special cases alike. From 1995/1996 until 2010 the number of non-taxed units is reported, but not for the years before. Dell et al. (2007) try to estimate the fraction of non-taxed by comparing the reported numbers of tax units to census reports about the number of adult population. According to their estimations this fraction drops from 94\% in 1993/1994 to 63\% back in 1945/46. \\

@@

\emph{Comparison of tax data and other data sources - advantages and shortcomings}

Based on this sections previous discussion table~\ref{datacomp} summarizes important differences between tax and survey data related to analyses of income inequality.

@@

\multicolinterrupt{
\begin{table}[H]
\caption{Comparison of tax and survey data\label{datacomp}} 
\begin{center}
\begin{tabular}{lcc}
\hline\hline
                                   & Tax data       & Survey data\\
\hline
Statistical unit                   & tax units (individuals/married couples)  & households\\                                   
Concepts of economic resources  & data-driven   & theory-driven\\
Population coverage & total population & sample \\ 
Top incomes covered                 & yes   & no\\
Low incomes covered?               & no        & yes\\
Comparison over time possible 	   & long       & short \\
Cross country comparison possible? & limited   & possible\\
\hline
\end{tabular}
\end{center}
\end{table}
}
%RF: damit haben wir WER, WANN, WAS, wollen wir noch ein WOMIT/WAS DARSTELLEN -> methodik -> reldist ? jaa

To define a standard of measuring economic resources and related inequality we discussed different dimensions of demands the data needs to meet. To sum up, ideally we want to measure \textit{disposable income} for \textit{all} swiss \textit{households} yearly for an extended period of time. Tax data has one weakness when it comes to the ideal income measure as we need to work with ``taxable income'' which is designed to serve taxation purposes. However, both concepts -- what needs to be taxed and what can be spent -- are related as both address a distinction between necessary and voluntary expenses. By subtracting federal taxes from taxable income we are getting closer to the theoretically ideal income measure of disposable income. However we can not subtract the more important cantonal and communal taxes. 
The second most important drawback with tax data is that it does not adequately address households. There are few situations in which the tax unit equals the household, that is for individuals living alone, with married partner and/or deductible children and nobody else. This ``classic'' household setup however became less common in the last decades so we need to assume statements about household inequality based on taxed data became more and more biased. 
%The swiss law seems to be slightly outdated in this context.
Finally however, another advantage of tax data (over survey data) is indeed the observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HBS data). The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are rather consistent for the complete observation period\footnote{Limitations might however exist due to tax evasion and major changes in the tax system, e.g. 1945. Furthermore, tax units with incomes too small to quality for federal taxation are not documented before 1995 and treated as zeros afterwards making it difficult to track income development in the lower percentiles.}. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.
